Simulating High-Dimensional Multivariate Data

### 2021-02-28 notebooks/gradseminar10feb2021_talk.rmd
re-add notebooks so Grant can check out new branch without losing files

---
subtitle: "Simulating Ultra High-Dimensional Multivariate Data"
author:
  - "Alfred G. **Schissler**, PhD"
  - "Alexander D. **Knudson**, MS"
  - "Tomasz J. **Kozubowski**, PhD"
  - "Anna K. **Panorska**, PhD"
  - "Juli **Petereit**, PhD"
  - "*University of Nevada, Reno (UNR)*"
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "21 Dec 2020 (*updated: `r Sys.Date()`)*"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    logo: ../images/hex-bigsimr.png
bibliography: ../bigsimr.bib
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
set.seed(12212020)
## devtools::install_github("ggobi/GGally")
## for ggpairs
library(GGally)
## xaringan::inf_mr('cmstats2020_talk.rmd')
```

@@

@@

- It is critical to realistically simulate data when conducting Monte Carlo studies and methods.
- But measurements are often **correlated**, **non-normal**, and **high dimensional** in this era of big data.
- One example are *RNA-seq data* obtained through high-throughput biomedical experiments.
- 3 out of 20,501 mRNA measurements from 1212 breast cancer patients (TCGA BRCA):

```{r, echo=F, out.width='60%', fig.align='center'}
knitr::include_graphics("../fig/realDataFig.png")
```

@@

- We propose a Gaussian-copula random vector algorithm; an augmented NORmal To Anything (NORTA) @Cario1997 method. 
- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.
- An article motivating, describing, and applying the methods; in development.

@@

```{r ch010-myProb, include=FALSE, cache = TRUE}
myProb <- 0.95
```

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
brcaRDS <- paste0( "../data/brca", round(myProb*100), ".rds" )
if ( !file.exists( brcaRDS ) )  {
    ## full workflow simulating RNA-seq data
    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
    lastClinical <- which( names(allDat) == 'tumorsize' )
    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
    dim(brca) ## num of genes 20501
    ## compute the median expression! avoid the 0s
    brcaMedian <- apply(brca, 2, median)
    ## retain top (1-myProb)*100% highest expressing genes for illustration
    cutPoint <- quantile( x = brcaMedian, probs = myProb )
    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    brca <- brca[ , genesToKeep ]
    ## ncol(brca) / 20501
    ## print(d <- length(genesToKeep))
    ## save data as rds
    saveRDS(brca, brcaRDS)
}
brca <- readRDS(file = brcaRDS)
d <- ncol(brca)
```

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( '../data/exampleGenes.rds' ) )  {
    set.seed(10192020)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = '../data/exampleGenes.rds')
}
exampleGenes <- readRDS('../data/exampleGenes.rds')
small <-  brca %>% select( exampleGenes )
knitr::kable(round(head( small, n = 5)),
      booktabs = TRUE,
      caption = 'mRNA counts for three selected high-expressing genes from the first five observations of the BRCA data set.'
)
```

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, out.width='60%', fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }

if ( !file.exists( '../fig/realDataFig.png' ) ) {
    p <- GGally::ggpairs( data = small,
                                        # lower = list( continuous = wrap(lowerfun)),
                         upper = list(
                             continuous = wrap('cor', method = "spearman")
                         )) + theme_bw()
    ggsave('../fig/realDataFig.png', plot = p, dpi = 300)
}
knitr::include_graphics("../fig/realDataFig.png")
```

@@

1. Pre-processing for nonparametric dependency matching.  
	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$.
	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
	(iii) If not compute a close semi-positive definite correlation matrix ${\bf \widetilde{R}_{Pearson}}$.  
2. Gaussian copula construction.  
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
3. Quantile evaluations.  
	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

@@

```{r ch040-biMAEtable, cache=F, echo=F, warning=F, message=F}
## tabMSE  <- allDat %>%
##    group_by( N, margins ) %>%
##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
tabMSE  <- allDat %>%
    filter( N == 100000) %>%
    group_by( N, type, margins ) %>%
##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
    summarize( MAE = mean( abs( rho - rho_hat )  ) )
## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
## ( tabMSE$mse - normMSE ) / normMSE
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse

knitr::kable(tabMSE,
             col.names = c("No. of random vectors",
                           "Correlation type",
                           "Distribution",
                           "Mean abs. error"),
             caption = '',
)
```

@@

```{r ch050-readBRCA, echo=FALSE, eval=TRUE, cache=F}
## full workflow simulating RNA-seq data
brcaRDS <- '../data/brca95.rds'
myProb = 0.95
brca <- readRDS(file = brcaRDS)
d <- ncol(brca)
## smaller set for prototyping
## d=5
## brca <- brca[ , 1:d]
```

@@

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```

@@

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
## Set the number of random vectors
n <- 10000
## construct margins
nb_margins <- make_nbinom_alist(sizes, probs)
## run sims
sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                   ensure_PSD = TRUE, cores = cores) 
```

```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
nb_margins <- make_nbinom_alist(sizes, probs)
brcaSimRDS <- paste0( "../results/brca", round(myProb*100), "sim.rds" )
sim_nbinom <- readRDS( brcaSimRDS )
```

@@

```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', fig.show = 'hold', out.width='40%', fig.asp=1}
## knitr::include_graphics("../fig/realDataFig.png")
knitr::include_graphics('../fig/ch050-simDataFig.png')
```

@@

## 4D - Assessing estimation efficiency

-  To assess the error in our correlation estimation above:

```{r ch050-quadlossECHO, echo=T, eval=F, cache=FALSE}
## Number of experiments
m <- 10
## Simulate random vectors equal to the sample size
n <- nrow(brca)
## convert outside for faster simulation
nb_Rho_p <- bigsimr::cor_convert( rho = nb_Rho,
                                 from = corType, to = "pearson" )
## ensure PSD
nb_Rho_p <- bigsimr::cor_nearPSD( G = nb_Rho_p )
## create m random vectors and estimate correlation
simRho <- replicate(n = m,
  expr = { tmpSim <- rvec(n = n , nb_Rho_p, nb_margins,
    type = 'pearson', ensure_PSD = FALSE, cores = cores);
    bigsimr::cor_fast( x = tmpSim, method = corType )} ,
simplify = FALSE)
## find quadratic loss at each rep
quadLoss <- unlist( lapply( simRho, rags2ridges::loss,
                           T = nb_Rho, type = "quadratic"))

```

```{r ch050-quadloss, echo=F, eval=T, cache=FALSE}
quadLossRDS <- paste0( "../results/brca", round(myProb*100), "quadLoss.rds" )
quadLoss <- readRDS( quadLossRDS )
```

## 4D - Assessing estimation efficiency

The `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.

```{r ch050-quadloss-summary, echo=TRUE, eval=T, cache=FALSE}
summary(quadLoss)
```

- This distribution could be compared to high-dimensional designed covariance estimators to guide to help decide whether the additional complexity and computation time are warranted.

## 4E - Scale up to high dimensional RNA-seq data

@@

### Cons
* Discrete margins can slightly disrupt the matching algorithm (due the probability of ties). 
* (Nearly) exact Pearson matching is not supported.

@@

- Walter W. Piegorsch, PhD
- Edward J. Bedrick, PhD
- Yves A. Lussier, MD
- Heather Knudson
- Eric Li, MS
- Fred Harris, PhD

@@

### CFE-CMSTATS 2020 organizers

 - Thanks to everyone who makes this event possible!
 - Professor Anna Panorska for the invitation to speak today.

>>>>

---
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author: "Alfred G. Schissler, Alexander D. Knudson, Edward J. Bedrick, , Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli Petereit, Walter W. Piegorsch, Duc Tran"
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "10 Feb 2021 (*updated: `r Sys.Date()`)*"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    logo: ../images/hex-bigsimr.png
bibliography: ../bigsimr.bib
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
library(rstanarm)
set.seed( 02102021 )
## devtools::install_github("ggobi/GGally")
## for ggpairs
library(GGally)
## xaringan::inf_mr('cmstats2020_talk.rmd')
```

@@

```{r ch010-myD, include=FALSE, cache = TRUE, eval=T}
d <- 1000
```

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
brcaRDS <- paste0( "../data/brca", d, ".rds" )
if ( !file.exists( brcaRDS ) | !file.exists( '../data/examplePatients.txt' ) )  {
    brca0 <- TCGA2STAT::getTCGA(disease = "BRCA",
                            data.type = "RNASeq",
                            p = 6L)
    brca1 <- as.data.frame(t(brca0$dat))
    gene_median <- apply(brca1, 2, median)
    gene_median_order <- order(gene_median, decreasing = TRUE)
    brca  <- brca1[,head(gene_median_order, d)]
    writeLines( substring( row.names( brca ), first=1, last= 12 ), '../data/examplePatients.txt' ) 
    saveRDS(brca, brcaRDS)
}
brca <- readRDS(file = brcaRDS)
```

@@

- It is critical to realistically simulate data when conducting Monte Carlo studies and methods.
- But measurements are often **correlated**, **non-normal**, and **high dimensional** in this era of big data.
- One example are *RNA-seq data* obtained through high-throughput biomedical experiments with Over  20,000 mRNA measurements from `r dim(brca)[1]` breast cancer patients (TCGA BRCA).

## Intergeneic correlation

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }  
GGally::ggpairs( data = small,
                # lower = list( continuous = wrap(lowerfun)),
                upper = list(
                    continuous = wrap('cor', method = "spearman")
                ), 
                ) + theme_bw()
```

@@

- We propose a Gaussian-copula random vector algorithm; an augmented NORmal To Anything (NORTA) @Cario1997 method. 
- We provide a high-performance and GPU-accelerated implementation in the `R` package [`bigsimr`](https://github.com/SchisslerGroup/bigsimr), along with related tools.
- An article motivating, describing, and applying the methods; in development.

@@

@@

1. Mapping step  
   i. Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.   
   or  
i$^\prime$. Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
2. Checking admissibility and, if needed, nearest correlation computation
    (i) Check that $R_X$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
	Otherwise,  
(ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $R_X$, in the Frobenius norm.
3. NORTA step
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
	(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

@@

```{r ch040-BiError, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## tabMSE  <- allDat %>%
##    group_by( N, margins ) %>%
##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
tabMAE  <- allDat %>%
    ## filter( N == 100000) %>%
    group_by( N, type, margins ) %>%
##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
    summarize( MAE = mean( abs( rho - rho_hat )  ) )
## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
## ( tabMSE$mse - normMSE ) / normMSE
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse

## AGS 12/22/20: Removing the latex formatting fixes the html render.
## But how to improve latex table formatting? Better to use the template
knitr::kable(tabMAE,
             ## format = 'latex',
             col.names = c("No. of random vectors",
                           "Correlation type",
                           "Distribution",
                           "Mean abs. error"),
             ## booktabs = TRUE,
             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
             ## linesep = c('', '',  '\\addlinespace')
)
```

@@

@@

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
type <- 'spearman'
R_S <- bigsimr::cor_fast(brca, method = type)
```

@@

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores) 
colnames(sim_nbinom) <- names(brca)
```

```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
brcaSimRDS <- paste0( "../results/brca", d, "sim.rds" )
if( !file.exists(brcaSimRDS) ) {
    set.seed(44)
    sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores)
    saveRDS( sim_nbinom, brcaSimRDS )
}
sim_nbinom <- readRDS( brcaSimRDS )
```

@@

```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', out.width='50%'}
## knitr::include_graphics("../fig/realDataFig.png")
knitr::include_graphics('../fig/ch050-simDataFig.png')
```

@@

## 4D - Scale up to high dimensional RNA-seq data

@@

### Cons
* Discrete margins can slightly disrupt the matching algorithm (due the probability of ties). 
* ~~exact Pearson matching is not supported.~~ EDIT: Pearson matching added via Hermite polynomial in `Julia`

@@

- Heather Knudson (graphic design)
- Eric Li, MS
- Fred Harris, PhD

@@

## Ongoing project HD multilevel survival model

* precision medicine of cancer 
* finding biomarkers from underrepresented patients

```{r readFits, echo=F}
fit <- fit0group <- readRDS( '~/article_survHD/results/fitOLDDATA_group_0_hs_ms_4_4_02162021_2000.rds')
## summary(fit)
## print(fit, digits=2)
## ps_check(fit)
```

```{r psCheckFinalFit0group, echo=F, cache=T}
raceLevels <- c('white', gsub( 'race', '', names(coef(fit0group))[2:3]) )
nd <- data.frame( race= raceLevels )
ps <- posterior_survfit(fit0group,
                        newdata = nd,
                        times = 0,
                        extrapolate = TRUE, control = list(edist = 8000))
panel_labels  <- c( '1' = "white", '2' = "black or african american", '3'= "asian")
pps <- plot(ps) + facet_wrap( ~ id, labeller = labeller(id=panel_labels))
pps
```

## Ongoing projects 2

- Other simulation and PMF calculations (*T-Poisson* article under minor revisions)
- Bayesian modeling of high-resolution NFL data (Erick Luerken, PhD student)
- Bayesian multilevel modeling of the pyschometric function 
- Bayesian multilevel modeling of US Electoral College (Andrey Sarantsev, Richard Foote).
- Bayesian imputation of missing single-cell RNA-seq data.
